---
title: "Get more out of surveydata with multilevel modelling"
author: "Andreas Tyge Moller"
date: "25 okt 2018"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Surveys on smoking habits can be analyzed more effectively
The smoking habits of Danish youth is frequently discussed these days, with
retailers and politicians competing to announce new measures and proposals for
curbing youth smoking.

Reliably measuring youth smoking frequencies is harder than you might expect 
though - see <http://altandetlige.dk/blog/6644/kommet-unge-dagligrygere-712>
(only available in Danish).

The official data on the smoking habits of the Danish population is collected
by every year by Danish Department of Health. A nationally representative sample
of 5,000 people are surveyed, and the National smoking frequency is measured by
the sample smoking frequency. That's it. No mension of uncertainty, no analysis.

But the data can be used much more efficiently. Just consider the 2016 survey.
The Danish Cancer Society collect data on smoking rates. Their survey samples
youth (aged 16-25) specifically, and 2,000 youth are surveyed. Youth are not 
equivalently grouped in the official data (but in the ages 15-19 and 20-29).
However, similar inferences could have been obtained from the official data, had 
you more efficiently exploited the information in the data.

This is where multilevel modelling comes in.

## What the smoking rates look like
Unfortunately the data behind the official smoking rates is unfortunately not
publicly available. The data are available annually after 1997, with a missing 
year in 2008. The smoking rates have experienced a consistent downward trend 
over the past 20 years.

```{r include=FALSE}
pacman::p_load(haven, dplyr, rstan, ggmcmc, rethinking, ggplot2)
# Load data
danskernes_rygevaner <- read_dta("C:/Users/Andreas Tyge Moller/Google Drive/Rygefrekvenser/DR_samlet.dta")
danskernes_rygevaner$smoker <- abs(danskernes_rygevaner$ryger_du - 2)
class(danskernes_rygevaner$smoker) <- "logical"
# Compute sample smoking prevalence
sample_smoking_prevalence <- danskernes_rygevaner %>%
  group_by(year) %>%
  summarise(sample_smoking_prevalence=mean(smoker, na.rm=T)*100)
```
```{r, echo=FALSE}
# Plot of sample smoking prevalence
ggplot(sample_smoking_prevalence, aes(year, sample_smoking_prevalence, group=1)) +
  geom_line(size=2, colour="#97d8f2") +
  theme_classic() +
  xlab("") + scale_x_discrete(breaks=seq(1998, 2016, 2)) +
  ylab("Sample smoking prevalence (percent)")
```

Data contain lots of variables, the majority of which are not fully consistent
over the years. Here I'll focus on the most central stratefication variables as
well as the main question "Do you smoke?". These variables have been
consistently collected since 2008, before which the survey setup was somewhat
different.

Since 2009 the smoking rate for young people (aged 16-29) has followed the 
overall smoking rate, but with larger deviations due to the smaller sample size 
(roughly 20 percent of the sample is below 30 years of age.).

```{r include=FALSE}
# Young smokers, ages 16-29
danskernes_rygevaner$young <- danskernes_rygevaner$alder7 %in% c(1, 2)
# After 2008 data are consistent until 2016
danskernes_rygevaner <- danskernes_rygevaner[which(danskernes_rygevaner$year>2008), ]
# Sample smoking prevalence, by young/not young. Age groups are only consistent after 2007.
sample_smoking_prevalence_grouped <- danskernes_rygevaner %>%
  group_by(year, young) %>%
  summarise(sample_smoking_prevalence=mean(smoker, na.rm=T)*100,
            group_size=n(),
            se=sqrt(var(smoker, na.rm = T)/n())*100)
# Plot of sample smoking prevalence, by young/not young. Age groups are only consistent after 2007.
```
```{r, echo=FALSE}
ggplot(data = sample_smoking_prevalence_grouped,
       aes(x = year, y = sample_smoking_prevalence, group = young, color = young)) +
  geom_line(size=2) +
  geom_errorbar(aes(ymin=sample_smoking_prevalence-se,
                    ymax=sample_smoking_prevalence+se, color=young), width=.1) +
  scale_color_manual(labels = c("Rest of sample", "Young people"),
                     values=c("#97d8f2", "#2a84a6")) +
  theme_classic() +
  theme(legend.position = c(0.9, 0.9)) +
  labs(color = "", x="", y="Sample smoking prevalence (percent)")  
```

I've added some error bars showing two standard deviations to get a feel for the
sample size. Looking at the two plots already gives cause for concern about
the reliability of the stratification of the survey sample. The
pattern of the overall smoking rate in 2014-2016 looks like it's driven by
noise. Comparing young people to the rest of the population, the changes
from year to year are implausibly large, and young people seem to have smoking 
rates inversely correlated with the rest of the sample. Not much can be learned
by looking at the changes from one year to the next. 

## Enter Bayesian data analysis
So what can we do to improve out understanding of the survey data? Bayesian 
modelling will allow us to incorporate prior knowledge into the data analysis.
Why is this is useful? We've already realized that the swings in the smoking
rate from year to year are implausible because we have an idea about how smoking
rates change over time. The Bayesian approach allow us to incorporate this
knowledge through a prior distribution.

To get a sense of how this works, I've estimated a simple model where the
smoking rates depend only on the year, but allow for there to be a relationship
between rates in different years. Below is what the data looks like.

```{r, echo=FALSE}
danskernes_rygevaner$smoker <- danskernes_rygevaner$smoker
danskernes_rygevaner$year <- as.factor(danskernes_rygevaner$year)
danskernes_rygevaner$female <- danskernes_rygevaner$KOEN - 1
attr(danskernes_rygevaner$female, "ATT") <- NULL
danskernes_rygevaner$age_group <- as.factor(danskernes_rygevaner$alder7)
model_data <- danskernes_rygevaner %>% 
  select(smoker, female, age_group, year) %>% 
  filter(complete.cases(.))
model_data <- data.frame(model_data)
```
```{r}
str(model_data)
```
```{r eval=FALSE}
m1 <- stan_glmer(formula = smoker ~ 1 + (1 | year),
                family = binomial(link = "logit"),
                data = model_data)
```
```{r, echo=FALSE}
load("rstanarm_m1.RData")
## Posterior mean and SD of each alpha
# draws for overall mean
m1.mu_a_sims <- as.matrix(m1, 
                       pars = "(Intercept)")
# draws for error
m1.u_sims <- as.matrix(m1, 
                    regex_pars = "b\\[\\(Intercept\\) year\\:")
# draws for the varying intercepts               
m1.a_sims <- plogis(as.numeric(m1.mu_a_sims) + m1.u_sims)
# Posterior mean and SD of each alpha
m1.mean <- apply(X = m1.a_sims,     # posterior mean
                 MARGIN = 2,
                 FUN = mean)
m1.sd <- apply(X = m1.a_sims,       # posterior SD
               MARGIN = 2,
               FUN = sd)
# Posterior median and 95% credible interval
m1.quant <- apply(X = m1.a_sims,
                  MARGIN = 2,
                  FUN = quantile,
                  probs = c(0.025, 0.50, 0.975))
m1.quant <- data.frame(t(m1.quant))
names(m1.quant) <- c("Q2.5", "Q50", "Q97.5")
# Combine summary statistics of posterior simulation draws
m1.df <- data.frame(m1.mean, m1.sd, m1.quant)
m1.df$year <- substring(rownames(m1.df), 20, 23)
```

Below is the posterior of the model (the regularized estimate), compared to the
raw sample mean smoking rate. I've included the 95 percent uncertainty interval 
to get a bit of a deeper sense of the model. Even with this simple model, it's 
easy to how the prior information and the estimated structure of year-to-year
changes islimiting the swings from year to year, and pulling the smoking rate 
towards the mean. Small changes, like 2012-2013, are barely affected, while 
implausibly large changes are regularized quite a bit. I've used the default 
priors in rstanarm, which are quite flat. Steeper (more conservative) priors 
would regularize the estimates even further.

```{r, echo=FALSE}
# Plot sample smoking prevalence and overlay estimated intercepts
sample_smoking_prevalence <- model_data %>%
  group_by(year) %>%
  summarise(sample_smoking_prevalence=mean(smoker)*100,
            group_size=n(),
            se=sqrt(var(smoker)/n())*100)
ggplot(data = sample_smoking_prevalence,
       aes(year, sample_smoking_prevalence, group=1)) +
  geom_line(color = "#97d8f2", size = 2) +
  theme_classic() +
  geom_line(data = m1.df,
            aes(year, m1.mean*100, group=1), size=2,  color="#d9a472") +
  geom_ribbon(aes(x = year,
                  ymin = (m1.mean - m1.sd)*100,
                  ymax = (m1.mean + m1.sd)*100), alpha=0.2) +
  xlab("") + scale_x_discrete(breaks=seq(2009, 2016, 1)) +
  ylab("Sample smoking prevalence (percent)") +
  geom_text(aes(label = "Sample mean", x = 1, y = 23.7), color="#97d8f2") +
  geom_text(aes(label = "Regularized \n estimate", x = 1, y = 22.7), color="#d9a472")
```

## Stay tuned - more detailed models to come...


















